Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2023-10-02 paper/paper-infer.tex
rm links to suppmat

\usepackage{xr}
\externaldocument{paper-infer-supp}

@@

First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2_main}--\ref{fig:simu_alpha_main} and \ref{fig:simu_h2}--\ref{fig:simu_alpha_large}).
Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figures \ref{fig:simu_h2_main}, \ref{fig:simu_h2} and \ref{fig:simu_h2_large}).
When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2-auto\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of $h^2$ (i.e.\ CIs are small, and do not contain the true parameter value, Figure \ref{fig:simu10_h2}), and LDpred2-auto\_jump overestimates the heritability (Figure \ref{fig:simu_h2}). LDpred2-auto\_nojump looks very reliable for estimating $h^2$.
When power is low, LDpred2-auto can overestimate the polygenicity when the true value is very small (e.g.\ $p=0.0005$), and underestimate it when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
SBayesS, which uses a similar model with the same three parameters, often overestimates the polygenicity, especially when $p\le0.02$.
The $\alpha$ estimate of LDpred2-auto (with the extended 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto.
Estimates of both $p$ and $\alpha$ from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figures \ref{fig:simu10_p} and \ref{fig:simu10_alpha}).
Generally, 95\% CIs for all three parameters ($h^2$, $p$ and $\alpha$) obtained from LDpred2-auto\_nojump contain the true simulated value (Figures \ref{fig:simu10_h2}--\ref{fig:simu10_alpha_large}), therefore confirming the validity of these CIs; this is the preferred method we recommend using.
Finally, we also investigate alternative ways of post-filtering chains in LDpred2-auto, for both prediction and inference (LDpred2-auto\_altfilter compared to LDpred2-auto\_nojump, Methods); results remain practically unchanged (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha_large}).

@@

Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIPs) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
LDpred2-auto provides PIPs that are more calibrated than with e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block by default (Figure \ref{fig:susie}).
Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low, and we show it works with results from SBayesS's Gibbs sampler as well (Figures \ref{fig:simu_r2} and \ref{fig:simu_r2_large}).
CIs for this $r^2$ estimate very often encompass the true simulated value, expect for LDpred2-auto\_jump when both $h^2$ and $N$ are small (Figures \ref{fig:simu10_r2} and \ref{fig:simu10_r2_large}).

We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a threshold corresponding to the prevalence.
Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$). The main difference is that the $h^2$ and $r^2$ estimates must be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used for transforming estimates of $h^2$ and $r^2$ when using $N_\text{eff}$ in inference methods \cite[]{grotzinger2021pervasive}.

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a small polygenicity (Figure \ref{fig:ukbb_h2}).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}).
As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figures \ref{fig:ukbb_r2} and \ref{fig:ukbb_r2_noMLE}). Heritability estimates for height are probably overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').

@@

To investigate whether estimates from LDpred2-auto are robust to some misspecifications, we test using two alternative LD references (Methods).
Using a smaller number of individuals for computing the LD matrix results in a slightly overestimated $p$ and $h^2$ (and $r^2$) with LDpred2-auto, while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure \ref{fig:alt_LD}).
Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can always get some chains not to diverge (and therefore get non-zero predictive performance) for the seven (three and four) phenotypes mentioned before when using the previous alternative LD references (Figure \ref{fig:compare_MLE}).

Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}.
Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure \ref{fig:jumpsign}). $h^2$ estimates can also be slightly lower, while $\alpha$ estimates are broadly consistent.
As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here (when using the N.W. European LD reference) and a standard sampling can be used in this case (Figure \ref{fig:jumpsign}).

Finally, we investigate different transformations to apply to some continuous phenotypes used here.
Indeed, 49 of the phenotypes used here seem log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference and prediction from LDpred2-auto.
We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation; it probably makes sense to transform these phenotypes.
We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
Except for bilirubin and lipoprotein(a) concentration, higher $h^2$ estimates and predictive performance $r^2$ are generally obtained with the RIN-transformation than the log-transformation.

@@

Here we use the same individuals as in the previous section. We investigate using the extended set of HapMap3 variants proposed here, HapMap3+ (Methods), which includes \textasciitilde37\% more variants on top of HapMap3 variants recommended to use for LDpred2 (i.e.\ 1,054,330 + 389,866 variants), to improve the genome coverage of this set. As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
In particular, a much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}), instead of 4--12 hours before.

@@

In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one LD block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (resp.\ 8) of them have at least 50\% (resp.\ 80\%) of their heritability explained by a single block (Figure \ref{fig:protein_local_h2}).

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a small negative correlation of -5.5\% with the probabilities of being causal.
To verify that this does not correspond to regions of low LD, we compute LD scores; the median probabilities of being causal have a small correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.

Finally, since the HapMap3+ variants still represent a rather small proportion of common variants, we showcase running LDpred2-auto using a more dense set of variants from a small region, as usually done for fine-mapping.
We identify the most significant HapMap3+ variant for height, rs2871960 on chromosome 3, and read all the variants within a 500kb distance that have a MAF larger than 0.005 and INFO score larger than 0.5 in the UK Biobank; there are 3881 such variants.
Then, we perform a GWAS for these variants using the same 305,338 training individuals as used before (Figure \ref{manhattan_height}).
Finally, using the resulting GWAS summary statistics and an LD reference from the same set of individuals, we run both SuSiE-RSS and LDpred2-auto (\_nojump). We test two different values for L, the maximum number of causal variants in SuSiE-RSS (10 and 100), and three different values for the maximum value of the estimated $p$ (1, 0.01, and 0.001).
Per-variant posterior probabilities of being causal (also known as posterior inclusion probabilities, PIPs) are very similar with both methods, especially when restricting $p$ in LDpred2-auto, which is similar to what SuSiE-RSS assumes (more conservative).

@@

As expected \cite[]{loh2018mixed}, intercepts from LD Score regression are increasing with sample size: 1.02 (SE: 0.008) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M.
SNP heritability estimates are 64.6\% (SE: 2.7), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively, and 60.2\% [95\% CI: 57.2, 63.2], 63.2\% [62.0, 64.4], and 54.2\% [53.9, 54.5] with LDpred2-auto.
As expected, estimated predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto) are increasing with sample size, with 29.6\% [28.7, 30.5], 42.7\% [42.2, 43.1], and 47.0\% [46.8, 47.1], respectively.
Note that these $r^2$ estimates are probably overestimated by the same margin as the (SNP heritability) $h^2$ estimates, and correspond to \textasciitilde49\%, \textasciitilde67.5\%, and \textasciitilde87\% of $h^2$, respectively.
Even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only, therefore an even larger sample size is required to be able to better predict height.
Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.2\% [1.0, 1.5], 2.3\% [2.0, 2.5], and 5.9\% [5.6, 6.3], consistent with results of simulations with a large polygenicity (p=10\%). 
Therefore, we estimate that height has at least 50,000 causal variants.
These results are similar irrespective of whether \texttt{allow\_jump\_sign} is used or not, which is surprising to us.
We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure \ref{fig:postp_height}).
As for $\alpha$ estimates from LDpred2-auto, they remain consistent, with -0.71 [-0.75, -0.67], -0.74 [-0.76, -0.72], and -0.78 [-0.82, -0.76], respectively.
Finally, we compute per-annotation heritability estimates from LDpred2-auto results to investigate functional enrichment. We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
Heritability enrichments that we obtain from LDpred2-auto results are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K, and of slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).

@@

A description of the eight external GWAS summary statistics used is provided in Table \ref{tab:sumstats}; these do not include UKBB individuals.
Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the extended or previous model and sampling (via parameter \texttt{use\_MLE}).
Because of the increased mismatch between external GWAS summary statistics and the LD reference we use here (compared to in simulations and UKBB analyses), we also explore multiple values for parameter \texttt{coef\_shrink}, which is a multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix in LDpred2-auto.
Note that, to transform $r^2$ and $h^2$ estimates to the liability scale (except for vitamin D, which is a continuous trait), we use the prevalence in the UK Biobank as the population prevalence, which may be slightly biased \cite{fry2017comparison,van2022reweighting}.
Results are presented in Figure \ref{fig:sumstats}.
In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); therefore it seems more useful to use this new set of variants for larger sample sizes.
Using the extended model and sampling scheme (with option \texttt{use\_MLE}) provides equal or better predictive performance except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix). 
The 3-parameter model also generally provides better predictive performance in the UK Biobank analyses, where power and robustness is often not an issue (Figure \ref{fig:ukbb_r2_vs}).
Maximum $r^2$ are achieved at different LD regularization coefficients \texttt{coef\_shrink} across phenotypes, reflecting possible substantial mismatches between the GWAS summary statistics and LD reference used.
However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is surprising to us.
As for estimates of $h^2$ and $r^2$ (inferred from the Gibbs sampler), they tend to increase with smaller values of \texttt{coef\_shrink}. 
This is also the case for estimates of $p$ when using \texttt{use\_MLE = TRUE}.
Estimates of $\alpha$ are largely constant but can become very small (capped at -1.5) in the case of robustness issues when using e.g.\ \texttt{use\_MLE = TRUE} and almost no regularization on the LD matrix (i.e.\ \texttt{coef\_shrink} close to 1).

>>>>

%\usepackage{xr}
%\externaldocument{paper-infer-supp}

@@

First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2_main}--\ref{fig:simu_alpha_main} and S1--S6).
Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figures \ref{fig:simu_h2_main}, S1 and S2).
When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2-auto\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of $h^2$ (i.e.\ CIs are small, and do not contain the true parameter value, Figure S7), and LDpred2-auto\_jump overestimates the heritability (Figure S1). LDpred2-auto\_nojump looks very reliable for estimating $h^2$.
When power is low, LDpred2-auto can overestimate the polygenicity when the true value is very small (e.g.\ $p=0.0005$), and underestimate it when the polygenicity is large (e.g.\ $p=0.1$, Figure S3).
SBayesS, which uses a similar model with the same three parameters, often overestimates the polygenicity, especially when $p\le0.02$.
The $\alpha$ estimate of LDpred2-auto (with the extended 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto.
Estimates of both $p$ and $\alpha$ from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figures S9 and S11).
Generally, 95\% CIs for all three parameters ($h^2$, $p$ and $\alpha$) obtained from LDpred2-auto\_nojump contain the true simulated value (Figures S7--S12), therefore confirming the validity of these CIs; this is the preferred method we recommend using.
Finally, we also investigate alternative ways of post-filtering chains in LDpred2-auto, for both prediction and inference (LDpred2-auto\_altfilter compared to LDpred2-auto\_nojump, Methods); results remain practically unchanged (Figures S1--S6).

@@

Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures S13 and S14).
We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIPs) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
LDpred2-auto provides PIPs that are more calibrated than with e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block by default (Figure S15).
Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low, and we show it works with results from SBayesS's Gibbs sampler as well (Figures S16 and S17).
CIs for this $r^2$ estimate very often encompass the true simulated value, expect for LDpred2-auto\_jump when both $h^2$ and $N$ are small (Figures S18 and S19).

We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a threshold corresponding to the prevalence.
Results are very similar as with the continuous phenotypes above (Figures S20--S23), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$). The main difference is that the $h^2$ and $r^2$ estimates must be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used for transforming estimates of $h^2$ and $r^2$ when using $N_\text{eff}$ in inference methods \cite[]{grotzinger2021pervasive}.

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a small polygenicity (Figure S24).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure S25).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure S26).
As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figures \ref{fig:ukbb_r2} and S27). Heritability estimates for height are probably overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').

@@

To investigate whether estimates from LDpred2-auto are robust to some misspecifications, we test using two alternative LD references (Methods).
Using a smaller number of individuals for computing the LD matrix results in a slightly overestimated $p$ and $h^2$ (and $r^2$) with LDpred2-auto, while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure S28).
When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure S29).
Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can always get some chains not to diverge (and therefore get non-zero predictive performance) for the seven (three and four) phenotypes mentioned before when using the previous alternative LD references (Figure S30).

Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}.
Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure S31). $h^2$ estimates can also be slightly lower, while $\alpha$ estimates are broadly consistent.
As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here (when using the N.W. European LD reference) and a standard sampling can be used in this case (Figure S31).

Finally, we investigate different transformations to apply to some continuous phenotypes used here.
Indeed, 49 of the phenotypes used here seem log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference and prediction from LDpred2-auto.
We first compare to using raw (untransformed) phenotypes in Figure S32; estimates of $p$ and $\alpha$ are highly consistent.
However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation; it probably makes sense to transform these phenotypes.
We then compare to using the rank-based inverse normal (RIN) transformation in Figure S33; estimates for $p$ and $\alpha$ are also highly consistent.
Except for bilirubin and lipoprotein(a) concentration, higher $h^2$ estimates and predictive performance $r^2$ are generally obtained with the RIN-transformation than the log-transformation.

@@

Here we use the same individuals as in the previous section. We investigate using the extended set of HapMap3 variants proposed here, HapMap3+ (Methods), which includes \textasciitilde37\% more variants on top of HapMap3 variants recommended to use for LDpred2 (i.e.\ 1,054,330 + 389,866 variants), to improve the genome coverage of this set. As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
In particular, a much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure S34). 
However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure S35), instead of 4--12 hours before.

@@

In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure S36.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one LD block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (resp.\ 8) of them have at least 50\% (resp.\ 80\%) of their heritability explained by a single block (Figure S37).

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure S38).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure S39.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a small negative correlation of -5.5\% with the probabilities of being causal.
To verify that this does not correspond to regions of low LD, we compute LD scores; the median probabilities of being causal have a small correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.

Finally, since the HapMap3+ variants still represent a rather small proportion of common variants, we showcase running LDpred2-auto using a more dense set of variants from a small region, as usually done for fine-mapping.
We identify the most significant HapMap3+ variant for height, rs2871960 on chromosome 3, and read all the variants within a 500kb distance that have a MAF larger than 0.005 and INFO score larger than 0.5 in the UK Biobank; there are 3881 such variants.
Then, we perform a GWAS for these variants using the same 305,338 training individuals as used before (Figure S40).
Finally, using the resulting GWAS summary statistics and an LD reference from the same set of individuals, we run both SuSiE-RSS and LDpred2-auto (\_nojump). We test two different values for L, the maximum number of causal variants in SuSiE-RSS (10 and 100), and three different values for the maximum value of the estimated $p$ (1, 0.01, and 0.001).
Per-variant posterior probabilities of being causal (also known as posterior inclusion probabilities, PIPs) are very similar with both methods, especially when restricting $p$ in LDpred2-auto, which is similar to what SuSiE-RSS assumes (more conservative, Figure S41).

@@

As expected \cite[]{loh2018mixed}, intercepts from LD Score regression are increasing with sample size: 1.02 (SE: 0.008) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M.
SNP heritability estimates are 64.6\% (SE: 2.7), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively, and 60.2\% [95\% CI: 57.2, 63.2], 63.2\% [62.0, 64.4], and 54.2\% [53.9, 54.5] with LDpred2-auto.
As expected, estimated predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto) are increasing with sample size, with 29.6\% [28.7, 30.5], 42.7\% [42.2, 43.1], and 47.0\% [46.8, 47.1], respectively.
Note that these $r^2$ estimates are probably overestimated by the same margin as the (SNP heritability) $h^2$ estimates, and correspond to \textasciitilde49\%, \textasciitilde67.5\%, and \textasciitilde87\% of $h^2$, respectively.
Even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only, therefore an even larger sample size is required to be able to better predict height.
Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.2\% [1.0, 1.5], 2.3\% [2.0, 2.5], and 5.9\% [5.6, 6.3], consistent with results of simulations with a large polygenicity (p=10\%). 
Therefore, we estimate that height has at least 50,000 causal variants.
These results are similar irrespective of whether \texttt{allow\_jump\_sign} is used or not, which is surprising to us.
We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure S42).
As for $\alpha$ estimates from LDpred2-auto, they remain consistent, with -0.71 [-0.75, -0.67], -0.74 [-0.76, -0.72], and -0.78 [-0.82, -0.76], respectively.
Finally, we compute per-annotation heritability estimates from LDpred2-auto results to investigate functional enrichment. We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
Heritability enrichments that we obtain from LDpred2-auto results are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K, and of slightly larger magnitude with N=1.6M (Figure S43).

@@

A description of the eight external GWAS summary statistics used is provided in Table \ref{tab:sumstats}; these do not include UKBB individuals.
Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the extended or previous model and sampling (via parameter \texttt{use\_MLE}).
Because of the increased mismatch between external GWAS summary statistics and the LD reference we use here (compared to in simulations and UKBB analyses), we also explore multiple values for parameter \texttt{coef\_shrink}, which is a multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix in LDpred2-auto.
Note that, to transform $r^2$ and $h^2$ estimates to the liability scale (except for vitamin D, which is a continuous trait), we use the prevalence in the UK Biobank as the population prevalence, which may be slightly biased \cite{fry2017comparison,van2022reweighting}.
Results are presented in Figure S44.
In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); therefore it seems more useful to use this new set of variants for larger sample sizes.
Using the extended model and sampling scheme (with option \texttt{use\_MLE}) provides equal or better predictive performance except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix). 
The 3-parameter model also generally provides better predictive performance in the UK Biobank analyses, where power and robustness is often not an issue (Figure S45).
Maximum $r^2$ are achieved at different LD regularization coefficients \texttt{coef\_shrink} across phenotypes, reflecting possible substantial mismatches between the GWAS summary statistics and LD reference used.
However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is surprising to us.
As for estimates of $h^2$ and $r^2$ (inferred from the Gibbs sampler), they tend to increase with smaller values of \texttt{coef\_shrink}. 
This is also the case for estimates of $p$ when using \texttt{use\_MLE = TRUE}.
Estimates of $\alpha$ are largely constant but can become very small (capped at -1.5) in the case of robustness issues when using e.g.\ \texttt{use\_MLE = TRUE} and almost no regularization on the LD matrix (i.e.\ \texttt{coef\_shrink} close to 1).
